---
title: "Rmagic EMT Tutorial"
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## MAGIC (Markov Affinity-Based Graph Imputation of Cells)

* MAGIC imputes missing data values on sparse data sets, restoring the structure of the data
* It also proves dimensionality reduction and gene expression visualizations
* MAGIC can be performed on a variety of datasets
* Here, we show the effectiveness of MAGIC on epithelial-to-mesenchymal transition (EMT) data

Markov Affinity-based Graph Imputation of Cells (MAGIC) is an algorithm for denoising and transcript recover of single cells applied to single-cell RNA sequencing data, as described in Van Dijk D *et al.* (2018), *Recovering Gene Interactions from Single-Cell Data Using Data Diffusion*, Cell <https://www.cell.com/cell/abstract/S0092-8674(18)30724-4>.

### Installation

To use MAGIC, you will need to install both the R and Python packages.

In R, run these commands to install MAGIC and all dependencies:

```{r install_Rmagic, eval=FALSE}
if (!require(devtools)) install.packages(devtools)
if (!require(Rmagic)) devtools::install_github("KrishnaswamyLab/magic/Rmagic")
```

In a terminal, run the following command to install the Python repository.

```{bash install_python_magic, eval=FALSE}
pip install --user git+git://github.com/KrishnaswamyLab/MAGIC.git#subdirectory=python
```

We'll install a couple more tools for this tutorial.

```{r install_extras, eval=FALSE}
if (!require(viridis)) install.packages("viridis")
if (!require(ggplot2)) install.packages("ggplot2")
if (!require(readr)) install.packages("readr")
if (!require(phateR)) install.packages("phateR")
```

If you have never used PHATE, you should also install PHATE from the command line as follows:

```{bash install_python_phate, eval=FALSE}
pip install --user phate
```

### Loading packages

We load the Rmagic package and a few others for convenience functions.

```{r load_packages}
library(Rmagic)
library(readr)
library(ggplot2)
library(viridis)
library(phateR)
```

### Loading data

The example data is located in the MAGIC Github repository.

```{r load_data}
# load data
data <- read_csv("../../../data/HMLE_TGFb_day_8_10.csv.gz")
data[1:5,1:10]
```

First, we need to remove lowly expressed genes and cells with small library size.

```{r}
# keep genes expressed in at least 10 cells
keep_cols <- colSums(data > 0) > 10
data <- data[,keep_cols]
# look at the distribution of library sizes
ggplot() +
  geom_histogram(aes(x=rowSums(data)), bins=50) +
  geom_vline(xintercept = 3000, color='red')
```

```{r}
# keep cells with at least 1000 UMIs and at most 15000
keep_rows <- rowSums(data) > 1000 & rowSums(data) < 15000
data <- data[keep_rows,]
```

We should library size normalize and transform the data prior to MAGIC. Many people use a log transform, which requires adding a "pseudocount" to avoid log(0). We square root instead, which has a similar form but doesn't suffer from instabilities at zero.

```{r normalize}
data <- library.size.normalize(data)
data <- sqrt(data)
```

### Running MAGIC

Running MAGIC is as simple as running the `magic` function. Because this dataset is rather large, we can increase `k` from the default of 5 up to 15.

```{r run_magic}
# run MAGIC
data_MAGIC <- magic(data, k=15, genes=c("VIM", "CDH1", "ZEB1"))
```

### Results

We can plot the data before and after MAGIC to visualize the results.

```{r plot_raw}
ggplot(data) +
  geom_point(aes(VIM, CDH1, colour=ZEB1)) +
  scale_colour_viridis(option="B")
ggsave('EMT_data_R_before_magic.png', width=5, height=5)
```

```{r plot_magic}
ggplot(data_MAGIC) +
  geom_point(aes(VIM, CDH1, colour=ZEB1)) +
  scale_colour_viridis(option="B")
ggsave('EMT_data_R_after_magic.png', width=5, height=5)
```

As you can see, the gene-gene relationships are much clearer after MAGIC. We can look at the entire smoothed matrix with `genes='all_genes'`, passing the original result to the argument `init` to avoid recomputing intermediate steps. Note that this matrix may be large and could take up a lot of memory.

```{r run_magic_full_matrix}
data_MAGIC <- magic(data, k=15, genes="all_genes", init=data_MAGIC)
as.data.frame(data_MAGIC)[1:5, 1:10]
```

### Visualizing MAGIC values on PCA

We can visualize the results of MAGIC on PCA as follows.

```{r run_pca}
gt <- reticulate::import("graphtools")
dat <- gt$base$Data(as.matrix(data_MAGIC$result), n_pca=2L)
data_MAGIC_PCA <- as.data.frame(dat$data_nu) # as.data.frame(prcomp(data_MAGIC)$x)
colnames(data_MAGIC_PCA) <- c("PC1", "PC2")
ggplot(data_MAGIC_PCA) +
  geom_point(aes(x=PC1, y=PC2, color=data_MAGIC$result$VIM)) +
  scale_color_viridis(option="B") +
  labs(color="VIM")
ggsave('EMT_data_R_pca_colored_by_magic.png', width=5, height=5)
```

### Visualizing MAGIC values on PHATE

We can visualize the results of MAGIC on PHATE as follows.

```{r run_phate}
data_PHATE <- phate(data, k=15)
ggplot(data_PHATE) +
  geom_point(aes(x=PHATE1, y=PHATE2, color=data_MAGIC$result$VIM)) +
  scale_color_viridis(option="B") +
  labs(color="VIM")
ggsave('EMT_data_R_phate_colored_by_magic.png', width=5, height=5)
```

