---
title: "Rmagic Tutorial"
output: 
  github_document: default
  html_document: default
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## MAGIC (Markov Affinity-Based Graph Imputation of Cells)

* MAGIC imputes missing data values on sparse data sets, restoring the structure of the data
* It also proves dimensionality reduction and gene expression visualizations
* MAGIC can be performed on a variety of datasets
* Here, we show the effectiveness of MAGIC on epithelial-to-mesenchymal transition (EMT) data

To use MAGIC, cite:
MAGIC: A diffusion-based imputation method reveals gene-gene interactions in single-cell RNA-sequencing data
Biorxiv (preprint) February 2017. DOI: doi.org/10.1101/111591

### Installation

To use MAGIC, you will need to install both the R and Python packages.

In R, run these commands to install MAGIC and all dependencies:

```{r eval=FALSE}
if (!require(devtools)) install.packages(devtools)
if (!require(Rmagic)) devtools::install_github("KrishnaswamyLab/magic/R")
```

In a terminal, run the following command to install the Pythonn repository.

```{bash eval=False}
pip install --user git+git://github.com/KrishnaswamyLab/MAGIC.git#subdirectory=python
```

We'll install a couple more tools for this tutorial.

```{r eval=FALSE}
if (!require(viridis)) install.packages("viridis")
if (!require(ggplot2)) install.packages("ggplot2")
if (!require(readr)) install.packages("readr")
if (!require(phateR)) install.packages("phateR")
```

If you have never used PHATE, you should also install PHATE from the command line as follows:

```{bash eval=FALSE}
pip install --user phate
```

### Running MAGIC

We load the Rmagic package and a few others for convenience functions.

```{r}
library(Rmagic)
library(ggplot2)
library(readr)
library(viridis)
library(phateR)
```

The example data is located in the MAGIC Github repository.

```{r}
# load data
clusters <- read_csv("https://raw.githubusercontent.com/KrishnaswamyLab/PHATE/master/data/MAP.csv", 
                     col_names=c('wells', 'clusters'))
bmmsc <- read_csv("https://github.com/KrishnaswamyLab/PHATE/raw/master/data/BMMC_myeloid.csv.gz")
bmmsc <- bmmsc[,2:ncol(bmmsc)]
bmmsc[1:5,1:10]
```

We should normalize the data prior to running MAGIC. For scRNAseq, it is always a good idea to library size normalize and then either log or square root transform the data. Here, we use square root transform.

```{r}
bmmsc <- Rmagic::library.size.normalize(bmmsc)
bmmsc <- sqrt(bmmsc)
```

Running MAGIC is as simple as running the `run_magic` function. We use `rescale_percent=99` to make the output range comparable to the input, but this is not strictly necessary for downstream analysis.

```{r}
# run MAGIC
bmmsc_MAGIC <- magic(bmmsc, genes=c("Mpo", "Klf1", "Ifitm1"), rescale_percent=99)
```

### Results

We can plot the data before and after MAGIC to visualize the results.

```{r}
ggplot(bmmsc) +
  geom_point(aes(Mpo, Klf1, colour=Ifitm1)) + 
  scale_colour_viridis()
ggsave('BMMSC_data_R_before_magic.png', width=5, height=5)
```

The data suffers from dropout to the point that we cannot infer anything about the gene-gene relationships.

```{r}
ggplot(bmmsc_MAGIC) +
  geom_point(aes(Mpo, Klf1, colour=Ifitm1)) + 
  scale_colour_viridis()
ggsave('BMMSC_data_R_after_magic.png', width=5, height=5)
```

As you can see, the gene-gene relationships are much clearer after MAGIC. These relationships also match the biological progression we expect to see - Ifitm1 is a stem cell marker, Klf1 is an erythroid marker, and Mpo is a myeloid marker.

We can look at the entire smoothed matrix with `genes='all_genes'`, passing the original result to the argument `init` to avoid recomputing intermediate steps.

```{r}
bmmsc_MAGIC <- magic(bmmsc, genes="all_genes", rescale_percent=99, init=bmmsc_MAGIC)
as.matrix(bmmsc_MAGIC)[1:5, 1:10]
```


### Visualizing MAGIC values on PHATE

We can visualize the results of MAGIC on PHATE as follows.

```{r}
bmmsc_PHATE <- phate(bmmsc)
ggplot(bmmsc_PHATE) +
  geom_point(aes(x=PHATE1, y=PHATE2, color=bmmsc_MAGIC$result$Klf1)) +
  scale_color_viridis() +
  labs(color="Mpo")
ggsave('BMMSC_data_R_phate_colored_by_magic.png', width=5, height=5)
```

